---
title: "Applied Linear Algebra"
author: "Alex Stephenson"
date: "7/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 1: Vectors 

A *vector* is an ordered list of numbers. 

```{r}
vec <- c(-1.1, 0.0, 3.6, -7.2)
vec
```

Vectors are equal $a = b$ when they have the same size and same corresponding entries 

```{r}

# Equal 
a <- c(1,2,3)
b <- c(1,2,3)

a == b

# Not Equal 
a2 <- c(1,2,4)
b2 <- c(2,5,6)

a2 == b2
```

Block (or stacked) vectors are made by concatenating two or more vectors 

```{r}
stack <- c(a,b)
stack
```

Subvectors are made by slicing a vector into sub-elements. Usually colon notation is used $a_{r:s}$ where the subscript is the *index range*. While most computer languages index starting at 0, R follows standard mathematical notation where vectors are indexed beginning at 1. 

```{r}
z <- c(1, -1, 2,0)
z[2:3]
```

Zero vectors are vectors with all elements equal to 0. 

```{r}
zeros <- rep(0, 4)
zeros
```
Unsurprisingly, the Ones vector is a vector with all elements equal to 1 

```{r}
ones <- rep(1, 4)
ones
```

Sparsity is an attribute of a vector and a vector is said to be sparse if many of its entries are zero. The number of nonzero entries of a n vector is denoted `nnz()`

```{r}
nnz <- function(vector){
    return(length(vector[vector == 0]))
}

nnz(c(1,2,0,0))
```

### Vector Addition 

Two vectors of the same size can be added together by adding the corresponding elements. This forms a new vector of the same size, called the *sum* of the vectors 

```{r}
v1 <- c(0,7,3)
v2 <- c(1,2,0)

v1 + v2

```

Subtraction is a different form of addition so works the same way 

```{r}
v1 <- c(1,9)
v2 <- c(1,1)

v1 - v2
```

Being a computer language R can do something funky when adding different length vectors 

```{r}
v1 <- c(1,2,3)
v2 <- c(1,2)

# The result here is due to "recycling" which is occasionally useful, but will lead to incorrect computations if not watched out for. 
v1 + v2
```

#### Properties of Vector Addition 

1. Commutative: $a + b = b + a$
2. Associative: $(a + b) + c = a + (b + c) = a + b + c$
3. Adding the zero vector leads to to no change in the original vector. 
4. Subtracting a vector from itself yields the 0 vector 

### Scalar-vector multiplication 

*Scalar multiplication* is when each element of a vector is multiplied by a scalar. A scalar means a number like 1, 46, or 3/5. 

```{r}
-2*c(1,9,6)
```

#### Properties 

1. Commutative: $\alpha a = a\alpha$
2. Associative: $(\beta\gamma)a = \beta(\gamma a)$
3. If a is a vector, $(\beta + \gamma)a = \beta a + \gamma a$
```{r}
(2 + 5)*v1 == 2*v1 + 5*v1
```

### Inner Product 

The standard *inner product* (aka "dot product") of two n-vectors is defined as the the scalar $a^Tb = a_1b_1 + ... + a_nb_n$ or the sum or the products of corresponding entries. 

```{r}
innerprod <- function(vec1, vec2){
    # Note that t() is the transpose function in R
    return(t(vec1)%*%vec2)
}

innerprod(c(-1,2,2), c(1,0,-3))
```

#### Properties 

1. Commutativity: $a^Tb = b^Ta$. This means that order does not matter for the arguments in the inner product. 
2. Associatitivy with scalar multiplication: $(\gamma a)^Tb = \gamma(a^Tb) = \gamma a^Tb$
3. Distributivity with vector additions $(a + b)^Tc = a^Tc + b^Tc$
4. "Expand the product": $(a + b)^T(c+d) = a^Tc + a^Td + b^Tc + b^Td$. Note that on the left hand side the "+" means vector addition and on the right the sums means scalar addition. 

For block vectors, if vectors *a* and *b* are block vectors and the corresponding blocks have the same size (they're conformable) then: $a^Tb = a_1^Tb_1 + ... a_k^Tb_k$

The inner product of the block vectors is the sum of the inner products of the block. 

```{r}
#We can use our inner product function to calculate co-cocurrances of two vectors 
# The number of objects in both A and B 
innerprod(c(0,1,1,1,1,1,1), c(1,0,1,0,1,0,0))
```
An example of where the inner product can be useful is giving the expected value of a quantity 

```{r}
exp_value <- function(probabilities, values){
    return(innerprod(probabilities, values))
}
values <- c(-1, 0, 5000)
probabilities <- c(.80, .15, .05)

exp_value(probabilities, values)


```
### Exercises

Suppose a class grade is determined by 10 assignments, where 8 are homework assignments on a 0-10 scale, one is a midterm on a 0-120 scale, and one is a final on a 0-160. The student's total course grade is a weighted average on a 0-100 scale is based on 25% homework, 35% midterm and 40% final. 

```{r}
calculate_course_grade <- function(hw, mid, final, scale){
    # We'll assume that scale is the weights 
    # which go in homework, midterm, final order 
    
    # Convert homework to 0-100 scale 
    # Because there are 8 assignments, we need 10/8
    # as a weight 
    hw_sum <- sum(10/8 * hw)
    
    # Convert midterm to 0-100 scale 
    # Midterm grade is out of 120 
    mid <- 100/120 * mid 
    
    # Convert final to 0-100 scale 
    # Final Grade is out of 160 
    final <- 100/160 * final 
    grades <- c(hw_sum, mid, final )
    
    # put into grade order 
    return(innerprod(grades, scale))
}

hw <- c(8,9, 5, rep(10,4))
mid <- 90
final <- 130
scale <- c(.25, .35, .4)
calculate_course_grade(hw, mid, final, scale)

```
#### Notes 

In R, we can use `crossproduct()` instead of our inner product function. 
